A generalization of the Bradley-Terry model for draws in chess with an application to collusion

### 2020-10-11 inst/jebo_chess_revision2.tex
title change suggested by Daniel Houser for increased visibility
remove colour coding for upload to JEBO

\title{A generalization of the Bradley-Terry model for draws in chess}

@@

\begin{abstract}
\definecolor{DarkGreen}{RGB}{14,159,4}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{DarkGreen}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-5}$).}
\end{abstract}

@@

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  \textcolor{orange}{First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.  More
  generally, \cite{szymanski2003} considers sporting contests as
  examples of competitive situations in which a contribution of some
  kind [effort, investment, attendance of the elite, etc] is elicited
  from contestants who may as a result win a prize.
  \citeauthor{szymanski2003} points out that this paradigm arguably
  covers executive corporate promotion, interviews for a position in
  an office, and most sporting tournaments.  As such, the phenomenon
  of collusion and indeed drawing in pairwise comparison is surely an
  interesting phenomenon.}

@@

\textcolor{blue}{The seminal work of \citet{moul2009} presents
  sophisticated regression techniques to analyse draws and first-mover
  advantage.  They compare all-play-all events with knockout format
  events to test---and reject---hypotheses of a distinct Soviet
  style-of-play; the work I present also considers Soviet proclivity
  to draw, using data from a single event, specifically the Stockholm
  Interzonal tournament of (1962).}

@@

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
  2017.  During this time, the players' strengths might have changed
  substantially.  Statistical assessment of time-varying strength is
  possible with the reified entity technique and this is discussed
  under further work below.  However, we may treat the strengths as
  constant by adopting one of two approaches: either we assume that
  the strengths do not change, or alternatively we interpret the
  strengths as measures of ability {\em averaged over a playing
    career}.  Either way,} Equation~\ref{likelihood_WD} becomes

@@

\textcolor{blue}{\citet{moul2009} consider this accusation from a
  statistical perspective.  Their analysis incorporated a dataset of
  7355 games played in the period 1940-1978, used to calibrate the
  ability of the players.  In contrast, the analysis presented here
  uses only game outcomes (win-loss-draw) in the competition itself.}

@@

\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
  the type considered here is ``effective'' in the sense that
  noncollusive games are detrimental to future play in a way that
  collusively drawn games are not.  At the end of the tournament,
  Petrosian, Keres and Geller had each played a total of 18
  noncollusive matches while Fischer, Benko and Filip had played a
  total of 26 or 27 (Tal, retiring mid tournament due to illness,
  played only 21).  For each game in the tournament, we count the
  total noncollusive games played up to that point for both the white
  and black player.  If, in that game, the number of previous
  noncollusive games for the white and black player differs by more
  than one\footnote{\textcolor{DarkGreen}{If collusivity is ignored,
    no game in the tournament is played between two players with
    cumulative game count differ by more than one.}}, we posit a ``rest
  monster'' whose strength helps the player who has played fewer
  noncollusive games.}

\textcolor{DarkGreen}{
  Observing that the rest monster is
  active in 49 of the 105 games of the tournament, we may consider
  these 49 games and, discarding draws, test whether rested players
  actually win more often than expected; observations are given in
  Table~\ref{restmonsterfisher} .  We reject the null of rested
  players having the same strength as their more tired opponents
  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}

\begin{table}
\textcolor{DarkGreen}{
  \caption{Results for the 25 won games (out of 49 played) in
    Cura\c{c}ao 1962 between rested and unrested
    players \label{restmonsterfisher}} \centering \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
  }
\end{table}

\textcolor{DarkGreen}{The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, estimates the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\% and the White advantage at about 3\%).  The null of the rest
  monster having zero strength is rejected with a likelihood ratio of
  about $e^{11.9}$ or a $p$-value of about $7\times 10^{-6}$.  It is
  an interesting observation that neither
  Table~\ref{restmonsterfisher}, nor reified Bradley-Terry, admit a
  White advantage (one-sided McNemar test, $p=\frac{1}{8}$, likelihood
  ratio $1.609$ respectively).  }

@@

The present work details an approach to chess analysis based on
reification of entities representing proclivity to draw, first-mover
(or white player) advantage, and \textcolor{DarkGreen}{the additional
  playing strength conferred by previous collusive draws.  Likelihood
  functions for the players' and reified entities' strengths are
  defined using only specific tournament outcomes.}  Many hypotheses
of interest to the chess community may be analysed straightforwardly.

@@

\textcolor{red}{One strong assumption made in the
  Kasparov-Karpov-Anand analysis was the constancy of the players'
  strengths over a 36-year period.  One approach might be to consider
  only matches played over a shorter span, but this would result in a
  smaller and less informative dataset; the tension between timeliness
  and informativeness is a delicate and difficult matter.}

\textcolor{orange}{In a wider context, many sports have a career arc
  whereby competitors start their careers weak, become stronger with
  experience, and finally become uncompetitive due to age.  Further
  work might include the introduction of additional reified entities
  that account for such phenomena: we might have a ``mid career
  monster'' who helps competitors at or near their career peak.}

\textcolor{orange}{The maximum likelihood estimate for strengths
  furnishes a natural ranking for the competitors.  \cite{lazear1981}
  observe that such ranking systems occur naturally in organizational
  situations (tournament theory) and the analysis presented here might
  be applicable to chess, with the advantage of allowing for
  both draws and incumbent advantage.}

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.
\textcolor{blue}{\cite{moul2009} go on to compare round robin
  tournaments to knockout format events, searching for evidence of a
  Soviet draw cartel.  The analysis presented here, only considering
  international tournaments [where Soviets have incentive to play
    differently against fellow Soviets or Westerners] does not admit
  this phenomenon, but the technique could easily be applied to a more
  informative dataset.}

\textcolor{DarkGreen}{The Bradley-Terry techniques used here furnish
  evidence that collusive games played during Cura\c{c}ao 1962 were
  less detrimental to future performance than noncollusive games.
  However, no cognitively realistic model for mental
  fatigue~\citep{boksem2008,linden2003} was used.  There was no
  allowance for recovery, for example; and the criterion for
  restedness itself was quite coarse.  It would be possible in
  principle to investigate such mechanisms using further reified
  entities.}

\subsection*{\textcolor{DarkGreen}{Acknowledgement}}

>>>>

\title{A generalization of the Bradley-Terry model for draws in chess with an application to collusion}

@@

\begin{abstract}
\definecolor{DarkGreen}{RGB}{14,159,4}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-5}$).
\end{abstract}

@@

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.  More
  generally, \cite{szymanski2003} considers sporting contests as
  examples of competitive situations in which a contribution of some
  kind [effort, investment, attendance of the elite, etc] is elicited
  from contestants who may as a result win a prize.
  \citeauthor{szymanski2003} points out that this paradigm arguably
  covers executive corporate promotion, interviews for a position in
  an office, and most sporting tournaments.  As such, the phenomenon
  of collusion and indeed drawing in pairwise comparison is surely an
  interesting phenomenon.

@@

The seminal work of \citet{moul2009} presents
  sophisticated regression techniques to analyse draws and first-mover
  advantage.  They compare all-play-all events with knockout format
  events to test---and reject---hypotheses of a distinct Soviet
  style-of-play; the work I present also considers Soviet proclivity
  to draw, using data from a single event, specifically the Stockholm
  Interzonal tournament of (1962).

@@

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}, showing games played from 1981 to
  2017.  During this time, the players' strengths might have changed
  substantially.  Statistical assessment of time-varying strength is
  possible with the reified entity technique and this is discussed
  under further work below.  However, we may treat the strengths as
  constant by adopting one of two approaches: either we assume that
  the strengths do not change, or alternatively we interpret the
  strengths as measures of ability {\em averaged over a playing
    career}.  Either way, Equation~\ref{likelihood_WD} becomes

@@

  \citet{moul2009} consider this accusation from a
  statistical perspective.  Their analysis incorporated a dataset of
  7355 games played in the period 1940-1978, used to calibrate the
  ability of the players.  In contrast, the analysis presented here
  uses only game outcomes (win-loss-draw) in the competition itself.

@@

It is reasonable to suppose that collusion of
  the type considered here is ``effective'' in the sense that
  noncollusive games are detrimental to future play in a way that
  collusively drawn games are not.  At the end of the tournament,
  Petrosian, Keres and Geller had each played a total of 18
  noncollusive matches while Fischer, Benko and Filip had played a
  total of 26 or 27 (Tal, retiring mid tournament due to illness,
  played only 21).  For each game in the tournament, we count the
  total noncollusive games played up to that point for both the white
  and black player.  If, in that game, the number of previous
  noncollusive games for the white and black player differs by more
  than one\footnote{If collusivity is ignored,
    no game in the tournament is played between two players with
    cumulative game count differ by more than one.}, we posit a ``rest
  monster'' whose strength helps the player who has played fewer
  noncollusive games.

  Observing that the rest monster is
  active in 49 of the 105 games of the tournament, we may consider
  these 49 games and, discarding draws, test whether rested players
  actually win more often than expected; observations are given in
  Table~\ref{restmonsterfisher} .  We reject the null of rested
  players having the same strength as their more tired opponents
  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).

\begin{table}
  \caption{Results for the 25 won games (out of 49 played) in
    Cura\c{c}ao 1962 between rested and unrested
    players \label{restmonsterfisher}} \centering \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
\end{table}

The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, estimates the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\% and the White advantage at about 3\%).  The null of the rest
  monster having zero strength is rejected with a likelihood ratio of
  about $e^{11.9}$ or a $p$-value of about $7\times 10^{-6}$.  It is
  an interesting observation that neither
  Table~\ref{restmonsterfisher}, nor reified Bradley-Terry, admit a
  White advantage (one-sided McNemar test, $p=\frac{1}{8}$, likelihood
  ratio $1.609$ respectively).

@@

The present work details an approach to chess analysis based on
reification of entities representing proclivity to draw, first-mover
(or white player) advantage, and the additional
  playing strength conferred by previous collusive draws.  Likelihood
  functions for the players' and reified entities' strengths are
  defined using only specific tournament outcomes.  Many hypotheses
of interest to the chess community may be analysed straightforwardly.

@@

One strong assumption made in the
  Kasparov-Karpov-Anand analysis was the constancy of the players'
  strengths over a 36-year period.  One approach might be to consider
  only matches played over a shorter span, but this would result in a
  smaller and less informative dataset; the tension between timeliness
  and informativeness is a delicate and difficult matter.

In a wider context, many sports have a career arc
  whereby competitors start their careers weak, become stronger with
  experience, and finally become uncompetitive due to age.  Further
  work might include the introduction of additional reified entities
  that account for such phenomena: we might have a ``mid career
  monster'' who helps competitors at or near their career peak.

The maximum likelihood estimate for strengths
  furnishes a natural ranking for the competitors.  \cite{lazear1981}
  observe that such ranking systems occur naturally in organizational
  situations (tournament theory) and the analysis presented here might
  be applicable to chess, with the advantage of allowing for
  both draws and incumbent advantage.

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.
\cite{moul2009} go on to compare round robin
  tournaments to knockout format events, searching for evidence of a
  Soviet draw cartel.  The analysis presented here, only considering
  international tournaments [where Soviets have incentive to play
    differently against fellow Soviets or Westerners] does not admit
  this phenomenon, but the technique could easily be applied to a more
  informative dataset.

The Bradley-Terry techniques used here furnish
  evidence that collusive games played during Cura\c{c}ao 1962 were
  less detrimental to future performance than noncollusive games.
  However, no cognitively realistic model for mental
  fatigue~\citep{boksem2008,linden2003} was used.  There was no
  allowance for recovery, for example; and the criterion for
  restedness itself was quite coarse.  It would be possible in
  principle to investigate such mechanisms using further reified
  entities.

\subsection*{Acknowledgement}
